---
title: "Assignment 4"
author: "Ryan Johnson"
date: "9/24/2020"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    theme: cosmo
---

## Description

In this exercise, I will estimate a linear regression model to predict the values of one of my continuous variables, monthly rent (__RNTP__), based on the values of the other variables. 

These independent variables are: 

1. Annual household income (__HINCP__)
2. Number of bedrooms (__BDSP__)
3. Decade of construction (__BLDG_AGE_dec__)
4. Household type (__HHT_label__)

```{r data import libraries, message = FALSE, include=FALSE}
library(ggplot2)
library(tidyverse)
```

```{r sci notation, message=FALSE, include=FALSE}
options(scipen = 9)
```

## Linear regression model

``` {r load data, message=FALSE}
hh_data_a3 <- read.csv("TravisHH.csv") %>% 
  mutate(HHT_2 = case_when(
    HHT_label=="Married couple household"~"Married couple",
    HHT_label=="Other family household: Female householder, no spouse present"~"Single parent family",
    HHT_label=="Other family household: Male householder, no spouse present"~"Single parent family",
    HHT_label=="Nonfamily household: Male householder: Not living alone"~"Non-family roommates",
    HHT_label=="Nonfamily household: Female householder: Not living alone"~"Non-family roommates",
    HHT_label=="Nonfamily household: Female householder: Living alone"~"Single occupant",
    HHT_label=="Nonfamily household: Male householder: Living alone"~"Single occupant"))
```

Below, "NP" is excluded because it interferes with HHT_2 - by definition, the only observations with a household size of 1 are single-occupant households. 

```{r linear regression, message=FALSE}
model1 <- lm(RNTP ~ HINCP + BDSP + BLDG_AGE_dec + HHT_2, 
            data = hh_data_a3)

summary(model1)
```

## Results

The results of this model present a variety of interesting information. First, the R-squared value of ~ 0.24 is fairly high, and indicates a decent model fit. 

Next, most of the relationships are statistically significant - with just a few exceptions. For the categorical variable (__BLDG_AGE_dec__), the base category was the _1940s_. On average, structures in this base case saw higher rents than every other category, although the relationship with the _1950s_ and _2010s_ categories were not statistically significant. The coefficients ranged from a minimum of -94.51893041 for the 2010s to a maximum of -358.38196226 for the 1980s. This means that structures built in the 1980s on average paid rents $358 less than those built in the 1940s, and so on. 

In the case of household income, (__HINCP__), we see a result indicating a positive correlation with a coefficient of 0.00183933. This result shows that for every additional dollar of household income, rents rise by roughly 1/5th of a cent; in other words, for every $1,000 increase in household incomes, rents rise by approximately $1.84. 

The coefficient estimate for bedrooms per household (__BDSP__) is 124.05191233, which means that every additional bedroom corresponds to an increase in monthly rent of about $124. 

Finally, with household type (__HHT_label__), the base category was married couple households. Compared to this case, the coefficient estimate for non-family roommate households is 63.51164579, or about $64 more per month. For single-occupant households the relationship was _not_ statistically significant. For single-parent families, this coefficient was -98.10395069, or roughly $98 less per month. 




