---
title: "Assignment 1 Submission"
author: "Ryan Johnson"
date: "9/7/2020"
output: html_document
---
```{r load libraries, include=FALSE}
library(tidycensus)
library(tidyverse)
```

### *Description*


The dataset I have assembled for this course is comprised of **household-level data** from the 2018 American Community Survey. This dataset includes the following seven variables:

* **Household income (HINCP)** (continuous)
* **Monthly rent (RNTP)** (continuous)
* **Monthly mortagage payment (MRGP)** (continuous)
* **Number of people in household (NP)** (continuous)
* **Household type (HHT)** (categorical)
* **Tenure in household (TEN**) (categorical)
* **Year built (YBL)** (categorical)

This data was filtered to a sample of **49,410 observations** by including only those households with an income above the single-person poverty level, and with *some* monthly rent payment or mortgage payment.

### *R segments used to define this dataset*


First, I pulled household-level ACS 2018 data using the tidycensus package.

```{r}
hh_vars_2018 <- pums_variables %>%
  distinct(year, survey, var_code, 
           var_label, data_type, level) %>%
  filter(level == "housing", year == 2018, survey == "acs1")
```


Next, I filtered this data to create my desired sample. 

```{r load household-level MA data filtered, message=FALSE, results='hide'}
hh_MA_data <- get_pums(variables = c("HINCP", 
                                      "RNTP", 
                                      "MRGP", 
                                      "NP", 
                                      "HHT",
                                      "TEN",
                                      "YBL"),
                        state = "MA",
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
  filter(HINCP > 12140, RNTP > 0 | MRGP >0) %>%
  select(HINCP, RNTP, MRGP, NP, HHT_label, TEN_label, YBL_label)
```
### *Final dataset*

This is the first ten rows of the final sample of data:

```{r, echo=FALSE}
hh_MA_data
```

```{r writeCSV, eval=FALSE, include=FALSE}
write_csv(hh_MA_data, "hh_MA_data.csv")
```

Lastly, I wrote this dataset to a CSV file for future use. 
