---
title: "Assignment 1 Submission"
author: "Ryan Johnson"
date: "9/9/2020"
output: html_document
---
```{r load libraries, include=FALSE}
library(tidycensus)
library(tidyverse)
```

### *Description*

The dataset I have assembled for this course is comprised of **household-level data** from the one-year 2018 American Community Survey (ACS). This dataset includes the following seven variables:

* **Household income (HINCP)** (continuous)
* **Monthly rent (RNTP)** (continuous)
* **Monthly mortagage payment (MRGP)** (continuous)
* **Number of people in household (NP)** (continuous)
* **Household type (HHT)** (categorical)
* **Tenure in household (TEN**) (categorical)
* **Year built (YBL)** (categorical)

This data was filtered to a sample of **8,269 observations** by including only those households in Travis County, TX, with an income above the single-person poverty level and with *some* monthly rent or mortgage payment.

I further filtered this data to only include observations taken in Travis County, TX, using a method described below.

### *R segments used to define this dataset*


First, I pulled household-level ACS 2018 data using the tidycensus package.


```{r}
hh_vars_2018 <- pums_variables %>%
  distinct(year, survey, var_code, 
           var_label, data_type, level) %>%
  filter(level == "housing", year == 2018, survey == "acs1")
```



Next, I filtered this data to create my desired sample. To accomplish this, I included the variable "PUMA", which lists the 5-digit Public Use Microdata Area codes, which are unique descriptors assigned to areas including roughly 100,000 people. These areas are the smallest unit of analysis Public Use Microdata from the Census Bureau. I used the Bureau's geography web map (https://tigerweb.geo.census.gov/tigerweb/) to find all of the codes included within Travis County. 

I limited my "get_pums()" function to include the above-mentioned variables and PUMAs, and then set the parameters to include 2018 data from TX. 

With the "filter()" function, I combined logical expressions to limit my dataset to only those observations taken in Travis County (PUMA codes 05301 through 05309) which met my income and housing cost criteria. 


```{r load hh Travis Co data, message=FALSE, results='hide'}
hh_TravisCo_data <- get_pums(variables = c("PUMA", 
                                   "HINCP", 
                                      "RNTP", 
                                      "MRGP", 
                                      "NP", 
                                      "HHT",
                                      "TEN",
                                      "YBL"),
                        state = "TX",
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
     mutate(PUMA = as.numeric(PUMA)) %>%
     filter(HINCP > 12140, RNTP > 0 | MRGP >0, PUMA > 05300 & PUMA < 05310) %>%
     select(HINCP, RNTP, MRGP, NP, HHT_label, TEN_label, YBL_label)
```



### *Final dataset*

This is the first ten rows of the final sample of data:


```{r, message=FALSE}
hh_TravisCo_data
```

```{r writeCSV, eval=FALSE, include=FALSE}
write_csv(hh_TravisCo_data, "hh_TravisCo_data.csv")
```

Lastly, I wrote this dataset to a CSV file for future use. 
